---
title: "Analysis of Tempus xT and xF data from Theranostic cohort for CRPC/Pluvicto"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

#Read in Libraries
```{r}
library(tidyverse)
library(clevRvis)
library(viridisLite)
library(maftools)
library(lubridate)
library(survminer)
library(ggsci)
library(gt)
library(ggconsort)

```





########################Demographic/Genomics plots######################################

#Plots of cohort describing metastatic sites, subtypes of aPCa, number of treatments
```{r}
#Preprocessing data
data <- read.csv(file = "/Users/debanjansaha/Desktop/pluvicto_plots_v2/Theranostics CRPC Tempus Dataframe.csv", nrows = 143, stringsAsFactors = F)#143 rows to ignore NA rows
data$study_id %>% unique() %>% length()
#View(data)

jco_cols <- pal_jco("default")(10)

#Date of treatments
data %>% 
    mutate(date_col = as.Date(tempus_date, format = "%m/%d/%y")) %>% summarise(
    earliest_date = min(date_col, na.rm = TRUE),
    latest_date   = max(date_col, na.rm = TRUE)
  )

#Number of each biopsy type
data %>%
    mutate(specimen_type = trimws(specimen))%>%
  filter(specimen_type %in% c("Liquid Biopsy", "Tissue")) %>%
  distinct(study_id, specimen_type) %>%
  group_by(study_id) %>%
  summarise(types = paste(sort(unique(specimen_type)), collapse = "+")) %>%
  count(types)


#Number who got pluvicto
got_pluvicto <- data %>%
  distinct(study_id, pluvicto_start) %>%
  mutate(Pluvicto = ifelse(is.na(pluvicto_start), "Naive", "Treatment")) %>% filter(Pluvicto == "Treatment")
```


```{r}

#library(ggconsort)

consort <- tibble(id = 1:442) %>%
  mutate(
    screened      = c(rep(TRUE, 67), rep(FALSE, 375)),
    got_pluvicto  = screened & c(rep(TRUE, 37), rep(FALSE, 30), rep(FALSE, 375)),
    psa_analysis  = got_pluvicto & c(rep(TRUE, 33), rep(FALSE, 4)),
    pre_post      = psa_analysis & (row_number() <= 4)
  )

study_cohorts <- consort %>%
  cohort_start("Patients referred for Theranostics evaluation") %>%
  cohort_define(
    excluded     = .full %>% filter(!screened),
    screened     = .full %>% filter(screened),
    no_pluvicto  = screened %>% filter(!got_pluvicto),
    pluvicto     = screened %>% filter(got_pluvicto),
    psa_analysis = pluvicto %>% filter(psa_analysis),
    pre_post     = psa_analysis %>% filter(pre_post)
  ) %>%
  cohort_label(
    excluded    = "Excluded: Not HSPC/CRPC and no tissue/cell-free Tempus data",
    screened    = "Screened: Genomic + Biopsy available",
    no_pluvicto = "Did not receive P-RPT",
    pluvicto    = "Received P-RPT with PSMA SPECT/CT imaging",
    psa_analysis = "\\>1 PSA timepoint", 
    pre_post    = "Pre/post biopsy pairs available"
  )

study_consort <- study_cohorts |>
  consort_box_add(
    "full", row = 1, col = 0,
    label = cohort_count_adorn(study_cohorts, .full)
  ) |>
  consort_box_add(
    "excluded", row = 2, col = "side",
    label = cohort_count_adorn(study_cohorts, excluded)
  ) |>
  consort_box_add(
    "screened", row = 3, col = 0,
    label = cohort_count_adorn(study_cohorts, screened)
  ) |>
  consort_box_add(
    "no_pluvicto", row = 4, col = -1,
    label = cohort_count_adorn(study_cohorts, no_pluvicto)
  ) |>
  consort_box_add(
    "pluvicto", row = 4, col = 1,
    label = cohort_count_adorn(study_cohorts, pluvicto)
  ) |>
  consort_box_add(
    "psa_analysis", row = 5, col = 1,
    label = cohort_count_adorn(study_cohorts, psa_analysis)
  ) |>
  consort_box_add(
    "pre_post", row = 6, col = 1,
    label = cohort_count_adorn(study_cohorts, pre_post)
  ) |>
  consort_arrow_add(start = "full", end = "screened") |>
  consort_arrow_add(start = "full", end = "excluded") |>
  consort_arrow_add(start = "screened", end = c("no_pluvicto", "pluvicto")) |>
  consort_arrow_add(start = "pluvicto", end = "psa_analysis") |>
  consort_arrow_add(start = "psa_analysis", end = "pre_post") |>
  consort_stage_add("Referral", row = c(1, 3)) |>
  consort_stage_add("Treatment allocation", row = 4) |>
  consort_stage_add("PSA follow-up", row = 5) |>
  consort_stage_add("Paired biopsy subset", row = 6)


pdf(file = "/Users/debanjansaha/Desktop/pluvicto_plots_v2/pluvicto_consort.pdf", width = 20)
study_consort |>
  ggplot() +
  geom_consort(label_size = 20) +     # `size` controls text size in the boxes (geom_label under the hood)
  theme_consort()
dev.off()


```


```{r}
#Met site plot
p1 <- data %>%
  separate_rows(met_sites, sep = "\\s*,\\s*") %>%
  dplyr::filter(!is.na(met_sites), met_sites != "") %>%
  distinct(study_id, met_sites) %>% ggplot(aes(y = met_sites)) + geom_bar(fill = jco_cols[1]) + theme_bw()  + ylab("Metastatic sites") 

#Subtype plot
p2 <- data  %>% distinct(study_id, diagnosis) %>% ggplot(aes(x = diagnosis, fill = 2)) + geom_bar(width = 0.5, fill = jco_cols[2]) + theme_bw() + xlab("Histologic subtype")


# Pluvicto therapy
p3 <- data %>%
  distinct(study_id, pluvicto_start) %>%
  mutate(Pluvicto = ifelse(is.na(pluvicto_start), "Naive", "Treatment")) %>%
  ggplot(aes(x = Pluvicto, fill = 3)) +
  geom_bar(width = 0.3, fill = jco_cols[3]) +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.5,      # positions the number just above the bar
    size = 4
  ) +
  theme_bw()  +
  xlab("Counts of patients receiving P-RPT")
#Gleason score
p4 <- data  %>%
  dplyr::filter(!is.na(gleason_score), gleason_score != "") %>%
  distinct(study_id, gleason_score) %>% ggplot(aes(y = gleason_score)) + geom_bar(fill = jco_cols[4]) + theme_bw()  + ylab("Gleason Score")

p5 <-data %>%
  dplyr::filter(!is.na(tumor_purity)) %>%
  mutate(
    tumor_purity = case_when(
      str_detect(tumor_purity, "<25") ~ 0,
      TRUE ~ as.numeric(tumor_purity)
    )
  ) %>%
  ggplot(aes(x = tumor_purity)) +
  geom_histogram(bins = 20, fill = jco_cols[5]) +
  theme_bw() +
  xlab("Tumor purity (%)") +
  ylab("Count")

p6 <- data %>%
  mutate(
    No.Lines.treatment =
      str_count(
        tx_context,
        "\\s*-\\s*"
      ) + 1
  ) %>%
  distinct(study_id, No.Lines.treatment) %>%
  ggplot(aes(x = No.Lines.treatment)) +
  geom_bar(fill = jco_cols[6]) +
  theme_bw() +
  xlab("Number of lines of treatment") +
  ylab("Count")

# pdf(file = "/Users/debanjansaha/Desktop/pluvicto_plots_v2/pluvicto_cohort.pdf", width = 10)
# ggarrange(p1, p2, p3, p4,p5, p6, ncol = 3, nrow = 2, legend = F)
# dev.off()

```

```{r}
#patient-level table (one row per study_id) 
pt_data <- data %>%
  distinct(study_id, diagnosis, met_sites, gleason_score, tx_context)

#met sites (multi-category, comma delimited
met_counts <- pt_data %>%
  separate_rows(met_sites, sep = ",\\s*") %>%
  count(met_sites) %>%
  mutate(pct = round(100 * n / n_distinct(pt_data$study_id), 1))

#diagnosis
diagnosis_counts <- pt_data %>% count(diagnosis)

#gleason: parse "3+4" -> total, take first value if multiple ("|"-separated)
gleason_clean <- pt_data %>%
  mutate(
    gleason_score   = na_if(gleason_score, "NA"),
    gleason_primary = str_extract(gleason_score, "^[^|]+"),
    total           = map_dbl(str_split(gleason_primary, "\\+"), ~ sum(as.numeric(.x))),
    gleason_group   = case_when(
      is.na(total) ~ NA_character_,
      total < 8    ~ "<8",
      total >= 8   ~ "8-10"
    )
  )
gleason_counts <- gleason_clean %>% count(gleason_group)

#tx_context: strip "Post-ARPI"/"Post ARPI" prefix, then split into tokens 
tx_long <- pt_data %>%
  mutate(tx_context = str_remove(tx_context, "^Post[- ]ARPI-?")) %>%
  separate_rows(tx_context, sep = "-") %>%
  filter(tx_context != "")

tx_long <- tx_long %>% filter(! (tx_context %in% c("Pre", "Post")) )
# 
# tx_counts <- tx_long%>%
#   mutate(tx_class = case_when(
#     grepl("axel", tx_context)       ~ "Taxane",
#     grepl("ARPI|amide", tx_context) ~ "ARPI",
#     grepl("Pluvicto", tx_context)   ~ "P-RPT",
#     TRUE ~ NA_character_
#   )) %>%
#   filter(!is.na(tx_class)) %>%
#   count(tx_class, sort = TRUE)

tx_counts <- tx_long %>% 
  group_by(study_id) %>% 
  mutate(count = n()) %>% 
  ungroup()
#Count summary for treatment lines
count_summary <- tx_counts %>% 
  summarise(
    min    = min(count),
    q1     = quantile(count, 0.25),
    median = median(count),
    mean   = mean(count),
    q3     = quantile(count, 0.75),
    max    = max(count)
  ) %>% 
  pivot_longer(everything(), names_to = "stat", values_to = "n")

#Pluvicto flag
pluvicto_counts <- pt_data %>%
  mutate(pluvicto_used = if_else(str_detect(tx_context, "Pluvicto"), "Yes", "No")) %>%
  count(pluvicto_used)

# ============ sample-level table (one row per study_id + tempus_date) ============
#tumor purity
purity_summary <- data %>%
  mutate(tumor_purity = as.numeric(tumor_purity)) %>%
  filter(!is.na(tumor_purity)) %>%
  summarize(median = median(tumor_purity),
            q1 = quantile(tumor_purity, .25),
            q3 = quantile(tumor_purity, .75))

#assay type / specimen source
#assay_counts    <- data %>% count(assay_type)
library(stringr)
specimen_counts <- data %>% mutate(specimen = str_trim(specimen)) %>% count(specimen)

# ============ print everything ============
results <- list(
  diagnosis  = diagnosis_counts,
  met_sites  = met_counts,
  gleason    = gleason_counts,
  tx_tokens  = count_summary,
  pluvicto   = pluvicto_counts,
  purity     = purity_summary,
  specimen   = specimen_counts
)


```


```{r}

gt_tables <- map(results, gt)
library(tidyverse)
library(gt)

#Labels
labels <- c(
  diagnosis  = "Diagnosis, No. (%)",
  met_sites  = "Metastatic sites, No. (%)",
  gleason    = "Gleason, No. (%)",
  pluvicto   = "P-RPT, No. (%)",
  specimen   = "Specimen source, No. (%)"
)

format_section <- function(df, label) {
  cat_col <- names(df)[1]   # first column = category
  n_col   <- "n"
  df %>%
    mutate(pct = round(100 * .data[[n_col]] / sum(.data[[n_col]]), 1)) %>%
    transmute(characteristic = paste0("  ", .data[[cat_col]]),
              value = paste0(.data[[n_col]], " (", pct, ")")) %>%
    add_row(characteristic = label, value = "", .before = 1)
}

#Counts for sections
count_sections <- map2(results[names(labels)], labels, format_section) %>%
  bind_rows()

#Tumor purity row
purity_row <- tibble(
  characteristic = "Tumor purity, median (IQR)",
  value = sprintf("%.0f (%.0f-%.0f)", results$purity$median, results$purity$q1, results$purity$q3)
)
#Number of treatment lines per pt including P-RPT
tx_count <- tibble(
  characteristic = "Treatment lines after ADT, median (IQR)",
  value = sprintf("%d (%d-%d)", 
                   as.integer(results$tx_tokens$n[3]), 
                   as.integer(results$tx_tokens$n[2]), 
                   as.integer(results$tx_tokens$n[5]))
)

table1 <- bind_rows(count_sections, tx_count, purity_row)

table1 %>%
  gt() %>%
  cols_label(
    characteristic = "Characteristic", 
    value = "Value"
  ) %>% tab_options(
    table.font.size = px(12),               # Slightly smaller font if needed
    data_row.padding = px(4)                # Tighter row padding
  ) %>%
   gtsave(filename = "/Users/debanjansaha/Desktop/pluvicto_plots_v2/demo_table.pdf", vheight = 900,                 
     vwidth = 1100)
```


########################Onco plots######################################

#SNV/Indel Oncoplot
```{r}
#SNV/Indel regex for Gene, Mutation,  VAF notation
mutation_regex <- "[A-Z0-9]+\\s*(p\\.[A-Za-z0-9\\*]+|splice_region)"

classify_mut <- function(hgvsp) {
  if (str_detect(hgvsp, "splice_region")) {
    return("Splice_Site")
  } else if (str_detect(hgvsp, "\\*")) {
    return("Nonsense_Mutation")
  } else if (str_detect(hgvsp, "fs")) {
    return("Frame_Shift_Del")
  } else if (str_detect(hgvsp, "p\\.")) {
    return("Missense_Mutation")
  } else {
    return("Other")  # catch-all instead of silently falling into Missense
  }
}

#---------Data wrangling for SNV/Indels----------------

#create list of mutations
mut_list <- data %>% 
  mutate(Tumor_Sample_Barcode = paste0(study_id, "_", tempus_date, "_", assay_type), mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, mutation_regex))
#Convert to df, drop empty entries 
mut_df <- mut_list %>%
  unnest(extracted) %>%       
  dplyr::filter(!is.na(extracted))   
#Separate gene from mutation in df
mut_df <- mut_df %>%
  mutate(
    Hugo_Symbol = str_extract(extracted, "^[A-Z0-9]+"),
    HGVSp_Short = str_extract(extracted, "p\\.[A-Za-z0-9\\*]+"),
    Variant_Classification = sapply(extracted, classify_mut)
  ) %>%
  dplyr::select(study_id, Tumor_Sample_Barcode, extracted, Hugo_Symbol, HGVSp_Short, Variant_Classification)


#Convert to MAF format with FS as FS-deletions (in truth can't distinguish if del from FS or not) or missense/nonsense as SNPs
snv_maf <- mut_df %>%
  mutate(
    Variant_Type = case_when(
      str_detect(HGVSp_Short, "fs") ~ "DEL",    #Classify FS as FS-Del
      TRUE ~ "SNP"  #Classify missense/nonsense as SNP                            
    )
  ) %>%
  transmute(
    #Creating placeholders to use with MAF plotting function
    Hugo_Symbol,
    Chromosome = "Unknown",
    Start_Position = 1,
    End_Position = 1,
    Reference_Allele = "N",
    Tumor_Seq_Allele2 = "N",

    #Required MAF fields
    Variant_Classification,
    Variant_Type,

    #AA change
    HGVSp_Short,

    #Sample ID
    Tumor_Sample_Barcode = Tumor_Sample_Barcode
  )
#Convert to MAF
maf_obj <- snv_maf%>% read.maf()

#Create clinical annotations per sample
clinical_annot <- data %>%
  mutate(
    Tumor_Sample_Barcode = paste0(study_id, "_", tempus_date, "_", assay_type),
    #Annot for biopsy type
    Assay = case_when(
      specimen == "Tissue" ~ "Tissue",
      specimen == "Liquid Biopsy" ~ "Liquid",
      TRUE ~ "Other"
    ),
    #Annot for pre/post Pluvicto
    Pluvicto = case_when(
      str_detect(tx_context, "Pre-Pluvicto") ~ "Pre-Pluvicto",
      str_detect(tx_context, "Pluvicto") ~ "Post-Pluvicto",
      TRUE ~ "None"
    ),
    #Annot for MSI status
    MSI = case_when(
      msi == "Stable" ~ "Stable",
      msi == "High" ~ "High",
      TRUE ~ "NA"
    ), 
    
   PSMA_FDG_discordance = case_when(
    str_detect(fdg_response, regex("discordance", ignore_case = TRUE)) ~ "Discordant",
    str_detect(fdg_response, regex("concordance", ignore_case = TRUE)) ~ "Concordant",
    is.na(fdg_response) ~ NA_character_
  )
    ) %>% 
  dplyr::select(Tumor_Sample_Barcode, Assay, Pluvicto, MSI, PSMA_FDG_discordance) %>% unique()

```



#CNA Oncoplot
```{r}
#Regex for CN mutations, annotated as Gene, CN Gain or CN loss
cna_regex <- "([A-Z0-9]+)\\s+(gain|loss|CN gain|copy number gain|amplification|amp|CN deletion|copy number loss|loss|del|copy_number_gain|copy_number_loss|CopyNumberLoss)"


#----------Data wrangling for CN alterations--------------
#Dataframe for CN Gains or Amps and CN Losses or Dels using regex
cna_df <- data %>%
  mutate(
    Tumor_Sample_Barcode = paste0(study_id, "_", tempus_date, "_", assay_type),
    cna_raw = str_extract_all(mut_status, cna_regex)
  ) %>%
  dplyr::select(Tumor_Sample_Barcode, cna_raw) %>%
  unnest(cna_raw) %>%
  dplyr::filter(!is.na(cna_raw)) %>%
   mutate(
    Hugo_Symbol = str_extract(cna_raw, "^[A-Z0-9]+"),
    direction = str_extract(
      cna_raw,
      regex("gain|loss|amplification|deletion", ignore_case = TRUE)
    ) %>% str_to_lower(),
    Variant_Classification = case_when(
      direction %in% c("gain", "amplification") ~ "Amplification",
      direction %in% c("loss", "deletion")      ~ "Deletion",
      TRUE ~ NA_character_
    )
  )

#Convert to wide format with rows as samples and columns as presence/absence of CN alteration
cna_wide <- cna_df %>%
  dplyr::select(Tumor_Sample_Barcode, Hugo_Symbol, CNA = Variant_Classification) %>%
  pivot_wider(
    names_from = Hugo_Symbol,
    values_from = CNA,
    values_fill = "None",
    values_fn = ~ paste(unique(.x), collapse = ";")
  )
#Merge CN dtaframe with prior clinical annotations 
clinical_annot2 <- clinical_annot %>%
  left_join(cna_wide, by = c("Tumor_Sample_Barcode" = "Tumor_Sample_Barcode"))
```


```{r}
#Pull PSA value columns and pivot long
psa_heat <- data %>%
  dplyr::select(study_id, 
                `Pre-C1` = psa_pre_c1_value,
                `C1`     = psa_c1_value,
                `C2`     = psa_c2_value,
                `C3`     = psa_c3_value,
                `C4`     = psa_c4_value,
                `C5`     = psa_c5_value,
                `C6`     = psa_c6_value) %>%
  distinct(study_id, .keep_all = TRUE) %>%
  mutate(across(-study_id, as.numeric)) %>%
  #Absolute difference columns
  mutate(across(c(`C1`, `C2`, `C3`, `C4`, `C5`, `C6`),
                ~ . - `Pre-C1`,
                .names = "diff_{.col}")) %>%
  #Percent change columns
  mutate(across(c(`C1`, `C2`, `C3`, `C4`, `C5`, `C6`),
                ~ (. - `Pre-C1`) / `Pre-C1` * 100,
                .names = "pct_{.col}")) %>%
  mutate(`Pre-C1` = 0) %>%
  dplyr::select(study_id, starts_with("diff_"), starts_with("pct_")) %>%
  pivot_longer(
    cols      = -study_id,
    names_to  = c(".value", "Cycle"),
    names_sep = "_"
  ) %>%
  mutate(Cycle = factor(Cycle, levels = c("Pre-C1","C1","C2","C3","C4","C5","C6"))) %>%
  group_by(study_id) %>%
  mutate(PSA50 = any(pct <= -50, na.rm = TRUE)) %>%
  ungroup()



```


```{r}
clinical_annot2 <- clinical_annot2 %>%
  mutate(study_id = str_extract(Tumor_Sample_Barcode, "^WU-LU-\\d+")) %>%  left_join(distinct(dplyr::select(psa_heat, c(study_id, PSA50))), by = "study_id") %>% dplyr::select(-study_id)

clinical_annot2 <- clinical_annot2 %>%
  rename(P_RPT = Pluvicto) %>%
  mutate(P_RPT = str_replace(P_RPT, "Pluvicto", "P-RPT"))  # keep hyphen only in the display VALUE, not the column name

#Vector of annotation plots
annot_vars <- c("Assay", "P_RPT", "MSI", "AR", "PTEN", "BRCA2", "PSA50", "PSMA_FDG_discordance")
annot_colors <- list()

#Set unique color per annotation plot, if none then Gray 
for (v in annot_vars) {
  cats <- unique(clinical_annot2[[v]])
  pal  <- setNames(jco_cols[5:10], cats)
  
  if ("None"   %in% cats) pal["None"]   <- "gray85"
 # if ("Stable" %in% cats) pal["Stable"] <- "gray85"
  if ("FALSE"  %in% cats) pal["FALSE"]  <- "gray85"
  if ("NA"       %in% cats) pal["NA"]       <- "gray85"  
  
  annot_colors[[v]] <- pal  
}

#---------Final plotting of Oncoplot-------------

#Set colors for SNVs/Indels/CNA
#pal <- viridis::viridis(n = 5, option = "D")
variant_colors <- c(
  Missense_Mutation = jco_cols[1],
  Nonsense_Mutation = jco_cols[2],
  Frame_Shift = jco_cols[3], 
  Amp = jco_cols[4],
  Del = jco_cols[5]
)

#Annot for TMB
tmb <-  data %>%
  mutate(
    Tumor_Sample_Barcode = paste0(study_id, "_", tempus_date, "_", assay_type),
    TMB = tmb) %>% dplyr::select(Tumor_Sample_Barcode, TMB) %>% unique()
tmb$TMB <- tmb$TMB %>% as.numeric()



#sample_order <- maf_obj@clinical.data$Tumor_Sample_Barcode 
pdf("/Users/debanjansaha/Desktop/pluvicto_plots_v2/pluvicto_oncoplot.pdf", width = 10)
oncoplot(
    maf = maf_obj,
    clinicalFeatures = c("Assay", "P_RPT", "MSI", "AR", "PTEN", "BRCA2", "PSA50", "PSMA_FDG_discordance"),
    annotationDat = clinical_annot2,annotationColor = annot_colors,
    colors = variant_colors, topBarData = tmb, titleText = "", legend_height = 5, anno_height = 5, legendFontSize = 1, annotationFontSize = 1
)
dev.off()
```


#VAF density plots of Pluvicto treated mCRPC
```{r}
library(ggExtra)
#Regex to extract VAFs from SNV data (Pattern is Gene, Mutation, VAF)
#vaf_regex <- "([A-Z0-9]+)\\s+(p\\.[A-Za-z0-9_*]+)\\s+([0-9]+\\.?[0-9]*)"
vaf_regex <- "([A-Z0-9]+)\\s+(p\\.[A-Za-z0-9_*]+|splice_region|c\\.[A-Za-z0-9_>+\\-]+)?\\s*([0-9]+\\.?[0-9]*)%?"

df <- data %>%
  group_by(study_id) %>%
  mutate(
    Pluvicto = as.integer(
      str_detect(tx_context,
                 regex("pluvicto", ignore_case = TRUE))
    )
  ) %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex), purity = as.numeric(tumor_purity))  %>%  unnest(extracted) %>%
  #Parse into gene/protein/vaf columns
  mutate(
    gene    = str_match(extracted, vaf_regex)[,2],
    protein = str_match(extracted, vaf_regex)[,3],
    vaf     = as.numeric(str_match(extracted, vaf_regex)[,4])
  ) %>% dplyr::select(study_id, Pluvicto, tempus_date, assay_type, gene, protein, vaf, tumor_purity) %>% mutate(
    sample_type = case_when(
      str_detect(assay_type, "xF") ~ "Liquid",
      str_detect(assay_type, "xT") ~ "Tissue",
      TRUE ~ NA
    ), mut_id = paste0(gene, "_", protein)) %>% 
   dplyr::select(study_id, tempus_date, mut_id, sample_type, vaf)  %>% dplyr::filter(!is.na(sample_type)) %>%
  pivot_wider(
  id_cols = c(study_id,
              tempus_date,
              mut_id),
  names_from = sample_type,
  values_from = vaf,
  values_fn  = \(x) x[[1]],          
  values_fill = NA_real_   
) %>%
  dplyr::filter(!is.na(Liquid) & !is.na(Tissue))
```


```{r}

#Build the per-mutation, per-sample dataframe 
df_long <- data %>%
  group_by(study_id) %>%
  mutate(
    Pluvicto = as.integer(str_detect(tx_context, regex("pluvicto", ignore_case = TRUE))),
    mutations = paste0(mut_status, "|", vus),
    extracted = str_extract_all(mutations, vaf_regex),
    purity    = as.numeric(tumor_purity)
  ) %>%
  unnest(extracted) %>%
  mutate(
    gene    = str_match(extracted, vaf_regex)[, 2],
    protein = str_match(extracted, vaf_regex)[, 3],
    vaf     = as.numeric(str_match(extracted, vaf_regex)[, 4])
  ) %>%
  dplyr::select(study_id, tempus_date, assay_type, gene, protein, vaf) %>%
  mutate(
    sample_type = case_when(
      str_detect(assay_type, "xF") ~ "Liquid",
      str_detect(assay_type, "xT") ~ "Tissue",
      TRUE ~ NA
    ),
    mut_id = paste0(gene, "_", protein)
  ) %>%
  dplyr::filter(!is.na(sample_type)) %>%
  dplyr::select(study_id, tempus_date, mut_id, sample_type, vaf)

#Split into liquid and tissue
df_liquid <- df_long %>%
  filter(sample_type == "Liquid") %>%
  mutate(liquid_date = tempus_date, Liquid = vaf) %>%
  dplyr::select(-sample_type)

df_tissue <- df_long %>%
  filter(sample_type == "Tissue") %>%
  mutate(tissue_date = tempus_date, Tissue = vaf) %>%
  dplyr::select(-sample_type)

#Join on study_id + mut_id, then filter to pairs within 30 days
df <- df_liquid %>%
  inner_join(df_tissue, by = c("study_id", "mut_id")) %>% mutate(
  liquid_date = mdy(liquid_date),   
  tissue_date = mdy(tissue_date),
  date_diff   = abs(liquid_date - tissue_date)
) %>%
  filter(date_diff <= 30)  %>%
  dplyr::select(study_id, liquid_date, tissue_date, mut_id, Liquid, Tissue)

```


```{r}
library(ggExtra)
n_genes <- df %>% mutate(Gene =gsub("_p\\..*$", "", mut_id)) %>% n_distinct()

p <- df %>% mutate(Gene =gsub("_p\\..*$", "", mut_id)) %>% ggplot( aes(x = Tissue, y = Liquid, color = Gene)) +
  geom_point(size = 3) +
  geom_abline(linetype = "dotted") +
  scale_color_jco() +
  theme_bw() +
  scale_color_manual(values = colorRampPalette(pal_jco()(10))(n_genes)) +
  xlim(0, 100) +
  ylim(0, 100) +
  labs(color = "Variant") + xlab("Tempus xT VAF") + ylab("Tempus xF VAF") + stat_cor(aes(group = 1), method = "spearman")
  
#pdf("/Users/debanjansaha/Desktop/pluvicto_plots_v2/pluvicto_vafplot.pdf", width = 10, height = 8)
ggMarginal(p, type = "density", fill = "black", alpha = 0.5)
#dev.off()

#Find average TMB
data$tmb %>% as.numeric %>% mean(na.rm = TRUE)
#Find MSI status distribution
data$msi %>% table() %>% prop.table()
```

########################Lollipop plots######################################

#Lolipop plots of select genes
```{r}
#Genes of interest
GOI <- c("AR", "TP53", "BRCA2", "PTEN", "ATM")
lapply(X = GOI, FUN = function(goi) {
  
#Loop thru SNV and VAF data for each gene in the annot df, separate into samples without Pluvicto  
tp53_obj_pre <- left_join(snv_maf, dplyr::select(clinical_annot2, Tumor_Sample_Barcode, P_RPT) )%>% dplyr::filter(Hugo_Symbol == goi)  %>% dplyr::filter(P_RPT == "None") %>% read.maf()

#SNV and VAF data for samples with Pluvicto treatment  
tp53_obj_post <- left_join(snv_maf, dplyr::select(clinical_annot2, Tumor_Sample_Barcode, P_RPT) )%>% dplyr::filter(Hugo_Symbol == goi)  %>% dplyr::filter(P_RPT != "None") %>% read.maf()

#Color variants as missense, nonsense, frameshift
lol_colors <- jco_cols[3:6]
names(lol_colors) <- unique(snv_maf$Variant_Classification)

#Plot lollipop plots to visulize exonic/intronic distribution of SNVs
pdf(paste0("/Users/debanjansaha/Desktop/pluvicto_plots_v2/", goi, "_lolipop.pdf"))
lollipopPlot2(m1 = tp53_obj_pre, m2 = tp53_obj_post , gene = goi, m1_name = "P-RPT Untreated", m2_name = "P-RPT Treated", colors = lol_colors, legendTxtSize = 2, labPosSize = 5, pointSize = 2 ) 
dev.off()
})




```





#Main idea: For patients with VAF data before and after Pluvicto, we Hcluster VAFs to identify clusters of VAFs implying clones
#Cut Tree to threshold these clones at given height, followed by manual inspection of clones using Pidgeon Hole principle to 
#Determine clonal ordering, namely VAF-parent >= Sum of all VAF-child recursed down the tree for each node. 
#Important Caveats: Hierarchical clustering is not as rigourous as other clustering algorithms (Sciclone, Pyclone) for clonal reconstruction, We do not have Copy number data for all exonic reads/VAFs, we assume baseline ploidy of 2 and 100% tumor purity and thus, are assuming that VAF is linearly proportional to the CCF. These are assumptions that are not necessarily true of all samples and thus limits our ability to infer clonal relationships. In the absence of this higher resolution data, these plots are an attempt to approximate the potential phylogenomic ordering although represent only one solution of many possible. 
########################Fish plots######################################

#Fishplot for WU-LU-051
```{r}
#Create matrix of VAF for each mutation/gene combo per row, ensure that filtering on times of starting and ending pluvicto  
mat <- data %>% dplyr::filter(study_id == "WU-LU-051") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex))  %>%  unnest(extracted) %>%  mutate(
    gene    = str_match(extracted, vaf_regex)[,2],
    protein = str_match(extracted, vaf_regex)[,3],
    vaf     = as.numeric(str_match(extracted, vaf_regex)[,4])) %>% mutate(protein =gsub(pattern = "VGS", replacement = "", x = protein) ) %>% mutate(mut_id = paste0(gene, "_", protein)) %>% dplyr::select(pluvicto_start, pluvicto_end, , tempus_date, assay_type, mut_id, vaf) %>% pivot_wider(names_from = mut_id, values_from = vaf)
purity <- data %>% dplyr::filter(study_id == "WU-LU-051") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex)) %>% dplyr::select(study_id, tempus_date, assay_type, tumor_purity)

#Filter for xF (cfDNA) Tempus data and format dates to datetime
vaf_mat <- mat %>% dplyr::filter(grepl(pattern = "xF|xT", x = assay_type)) %>% mutate(Date = as.Date(
  tempus_date,
  format = "%m/%d/%y")) %>% arrange(Date)%>% dplyr::select(-c(pluvicto_start, pluvicto_end, tempus_date, assay_type, Date))  %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>% t()

#distance on VAF profiles
d <- dist(vaf_mat)

# hierarchical clustering
hc <- hclust(d, method = "average")

plot(hc)
#Cut tree by distance threshold, using manual inspection
#Will remove BRCA1/JAK1 cluster as likely private germline mutations that don't change longitudinally as well as 3rd cfDNA sample due to low purity 
groups <- cutree(hc, h = 20)

```


```{r}
#Format colnames as timepoints
cnames <-c( paste0("T", seq_len(ncol(vaf_mat))), "groups")
#Appending Grouping to VAF data
vaf_mat <- data.frame(vaf_mat, groups)
#Reset colnames
colnames(vaf_mat) <- cnames

#Average mutations based on grouping from Hclustering
vaf_mat <- vaf_mat %>% mutate(mutation = row.names(.)) %>% 
 group_by(groups) %>%
  summarize(
    across(where(is.numeric), mean),
    mutations = paste(mutation, collapse = "\n"),
    .groups = "drop"
  )  %>%
  dplyr::select(-groups) %>%                 
  column_to_rownames("mutations") %>% 
  as.matrix()

#Manual inspection of clone-averaged VAF df with timepoints
data.frame(vaf_mat)
vaf_mat <- vaf_mat[1:3, c(1,2,3,5)]

#Scraped from EPIC Tempus reports, using median VAF * 2 as tumor purity for ctDNA and assuming those represent clonal mutations
#T1: 0.4% (xF)
#T2: 1.3% (xF)
#T3: 60% (xT)
#T4: <0.25% (xF) --> Remove due to low purity (LLOD)
#T5: 80% (xT) 
purity <- c(T1 = 0.4 * 2, T2 = 1.3 * 2, T3 = 60,  T4 = 80)
mat_corrected <- sweep(vaf_mat, 2, purity/100, "/") + 0.1 #Add small VAF to allow plotting
#Pin CCF > 100 as clonal
mat_corrected <- pmin(mat_corrected, 100)

mat_corrected


```


```{r}
#Parent assignment based on pigeonhole principle
parents <- c(0, 1, 1)
timepoints <- 1:4


#Format dates
mat <-  mat %>% dplyr::filter(!(assay_type == "Tempus xF" & tempus_date == "5/1/24")) #Remove this data manually given low tumor purity

tempus_dates <- mdy(mat %>% dplyr::filter(assay_type == "Tempus xF" | assay_type == "Tempus xT") %>% .$tempus_date) %>% sort()
pluvicto_dates <- mdy(unique(c(mat$pluvicto_start, mat$pluvicto_end))) %>% sort()
t0 <- min(tempus_dates)

#Get time from initial timepoint
tempus_time_numeric <- as.numeric(tempus_dates - t0)
pluvicto_time_numeric <- as.numeric(pluvicto_dates - t0)

#Grep CN data
cn_df <- data %>%
  dplyr::filter(study_id == "201169764" & assay_type == "Tempus xT") %>%
  mutate(
    Tempus_date = mdy(tempus_date)
  ) %>%
  dplyr::select(
    study_id,
    tempus_date,
    mut_status
  ) %>%
  mutate(
    cn_events = str_extract_all(
      mut_status,
      regex(
         "(\\b[A-Z0-9]+\\b)\\s*(CN gain|copy number gain|amplification|amp|CN deletion|copy number loss|loss|del|copy_number_gain|copy_number_loss|CopyNumberLoss)",
        ignore_case = TRUE
      )
    )
  ) %>%
  unnest(cn_events) %>%
  mutate(
    gene = str_extract(cn_events, "^[A-Z0-9]+"),
    cn_type = case_when(
      str_detect(cn_events, regex("gain|amp", ignore_case = TRUE)) ~ "gain",
      str_detect(cn_events, regex("loss|del", ignore_case = TRUE)) ~ "loss"
    )
  )%>% 
  mutate(
    time_numeric = as.numeric(mdy(tempus_date) - t0)
  )

#Create annotations for lines of times and labels
all_vlines <- c(
  tempus_time_numeric,
  pluvicto_time_numeric)

all_vlabels <- c(
  format(tempus_dates, "%b %Y"),
  paste0("PRPT:", format(pluvicto_dates, "%b %Y")))

#CN event label associated with tempus reports for GOIs
cn_annotations <- cn_df %>% dplyr::filter(gene %in% c("TP53", "AR", "BRCA2", "ATM", "PTEN", "RB1")) %>% 
  group_by(time_numeric) %>%
  summarise(
    x   = dplyr::first(time_numeric),
    y   = 110,
    lab = paste(
      paste0(gene, " ", cn_type),
      collapse = ":"
    ),
    col = c("black"),
    .groups = "drop"
  )

#Plotting fishplots
seaObject <- createSeaObject(round(mat_corrected, 2), parents, timepoints = tempus_time_numeric,
                            timepointInterpolation = T, col = pal_jco(alpha = 0.5)(10)[c(1:3)], cloneLabels = row.names(vaf_mat))

pdf("/Users/debanjansaha/Desktop/pluvicto_plots_v2/WU-LU-051_dolphinplot.pdf", width = 10)
dolphinPlot(
  seaObject,
  vlines = all_vlines,
  vlab   = all_vlabels,
  vlabSize = 2.5, ,annotSize = 3,annotations = cn_annotations,
  showLegend = TRUE,
  markMeasuredTimepoints = tempus_time_numeric,
  borderCol = "black",
  shape = "spline",
  pos = "center",
  separateIndependentClones = TRUE, ylab = "Variant allele frequency", xlab = "Time"
)
dev.off()
```

#Fishplot for WU-LU-054
```{r}
#Extract SNVs/Indels for annotated mutations and vus
mat <- data %>% dplyr::filter(study_id == "WU-LU-054") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex))  %>%  unnest(extracted) %>%  mutate(
    gene    = str_match(extracted, vaf_regex)[,2],
    protein = str_match(extracted, vaf_regex)[,3],
    vaf     = as.numeric(str_match(extracted, vaf_regex)[,4])) %>% mutate(protein =gsub(pattern = "VGS", replacement = "", x = protein) ) %>% mutate(mut_id = paste0(gene, "_", protein)) %>% dplyr::select(pluvicto_start, pluvicto_end, , tempus_date, assay_type, mut_id, vaf) %>% pivot_wider(names_from = mut_id, values_from = vaf)


purity <- data %>% dplyr::filter(study_id == "WU-LU-054") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex)) %>% dplyr::select(study_id, tempus_date, assay_type, tumor_purity)


#Filter for Tempus reports and convert to datetimes
vaf_mat <- mat %>% dplyr::filter(grepl(pattern = "xT", x = assay_type)) %>% mutate(Date = as.Date(
  tempus_date,
  format = "%m/%d/%y")) %>% arrange(Date)%>% dplyr::select(-c(pluvicto_start, pluvicto_end, tempus_date, assay_type, Date))  %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>% t()

#Distance of VAFs using H clustering
d <- dist(vaf_mat)
hc <- hclust(d, method = "average")

#Manually inspect clustering to determine cutoff
plot(hc)
# cut tree by distance threshold
groups <- cutree(hc, h = 10)  # tune h

#Rename timepoints
cnames <-c( paste0("T", seq_len(ncol(vaf_mat))), "groups")
vaf_mat <- data.frame(vaf_mat, groups) 
colnames(vaf_mat) <- cnames
#Average clusters in same grouping based on clustering and concatenate labels to get clones
#*We assume that VAF is proportional to CCF although no CN data to determine this**
vaf_mat <- vaf_mat %>% mutate(mutation = row.names(.)) %>% 
 group_by(groups) %>%
  summarize(
    across(where(is.numeric), mean),
    mutations = paste(mutation, collapse = "\n"),
    .groups = "drop"
  )  %>%
  dplyr::select(-groups) %>%                 # drop grouping column
  column_to_rownames("mutations") %>% # set clone names
  as.matrix()

#Manually inspect VAF trajectories to determine parent child relationships, more than one solution may be possible
data.frame(vaf_mat)
purity <- c(T1 = 50, T2 = 80)
mat_corrected <- sweep(vaf_mat, 2, purity/100, "/")

mat_corrected

```




```{r}
parents <- c(0, 1, 0, 1)
timepoints <- c(1, 2)


#Get tempus and pluvicto dates
tempus_dates <- mdy(mat %>% dplyr::filter(assay_type == "Tempus xT") %>% .$tempus_date) %>% sort()
pluvicto_dates <- mdy(unique(c(mat$pluvicto_start, mat$pluvicto_end))) %>% sort()
t0 <- min(tempus_dates)

tempus_time_numeric <- as.numeric(tempus_dates - t0)
pluvicto_time_numeric <- as.numeric(pluvicto_dates - t0)

#Grep CN data
cn_df <- data %>%
  dplyr::filter(study_id == "WU-LU-054" & assay_type == "Tempus xT") %>%
  mutate(
    Tempus_date = mdy(tempus_date)
  ) %>%
  dplyr::select(
    study_id,
    Tempus_date,
    mut_status
  ) %>%
  mutate(
    cn_events = str_extract_all(
      mut_status,
      regex(
         "(\\b[A-Z0-9]+\\b)\\s*(CN gain|copy number gain|amplification|amp|CN deletion|copy number loss|loss|del|copy_number_gain|copy_number_loss|CopyNumberLoss)",
        ignore_case = TRUE
      )
    )
  ) %>%
  unnest(cn_events) %>%
  mutate(
    gene = str_extract(cn_events, "^[A-Z0-9]+"),
    cn_type = case_when(
      str_detect(cn_events, regex("gain|amp", ignore_case = TRUE)) ~ "gain",
      str_detect(cn_events, regex("loss|del", ignore_case = TRUE)) ~ "loss"
    )
  )%>% 
  mutate(
    time_numeric = as.numeric(Tempus_date - t0)
  )

#Create annotations for lines of times and labels
all_vlines <- c(
  tempus_time_numeric,
  pluvicto_time_numeric)[c(1,3,4,2)]

all_vlabels <- c(
  format(tempus_dates, "%b %Y"),
  paste0("PRPT\n", format(pluvicto_dates, "%b %Y")))[c(1, 3, 4, 2)]

#CN event label associated with tempus reports for GOIs
cn_annotations <- cn_df %>% dplyr::filter(gene %in% c("TP53", "AR", "BRCA2", "ATM", "PTEN", "RB1")) %>% 
  group_by(time_numeric) %>%
  summarise(
    x   = dplyr::first(time_numeric),
    y   = 110,
    lab = paste(
      paste0(gene, " ", cn_type),
      collapse = ":"
    ),
    col = c("black"),
    .groups = "drop"
  )

#Create Seaobject
seaObject <- createSeaObject(round(mat_corrected, 0), parents, timepoints = tempus_time_numeric,
                            timepointInterpolation = T, col = pal_jco(alpha = 0.5)(10)[1:4], cloneLabels = row.names(vaf_mat))

pdf("/Users/debanjansaha/Desktop/pluvicto_plots_v2/WU-LU-054_dolphinplot.pdf", width = 10, height = 10)
#Create dolphin plot
dolphinPlot(
  seaObject,
  vlines = all_vlines,
  vlab   = all_vlabels,
  vlabSize = 4, annotations = cn_annotations,annotSize = 4,
  showLegend = TRUE,
  markMeasuredTimepoints = tempus_time_numeric,
  borderCol = "black",shape = "spline",
  pos = "center",
  separateIndependentClones = TRUE, ylab = "Variant allele frequency", xlab = "Time"
)
dev.off()


```


#WU-LU-066
```{r}
#Extract SNVs/Indels for annotated mutations and vus
mat <- data %>% dplyr::filter(study_id == "WU-LU-066") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex))  %>%  unnest(extracted) %>%  mutate(
    gene    = str_match(extracted, vaf_regex)[,2],
    protein = str_match(extracted, vaf_regex)[,3],
    vaf     = as.numeric(str_match(extracted, vaf_regex)[,4])) %>% mutate(protein =gsub(pattern = "VGS", replacement = "", x = protein) ) %>% mutate(mut_id = paste0(gene, "_", protein)) %>% dplyr::select(pluvicto_start, pluvicto_end, , tempus_date, assay_type, mut_id, vaf) %>% pivot_wider(names_from = mut_id, values_from = vaf)


purity <- data %>% dplyr::filter(study_id == "WU-LU-066") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex)) %>% dplyr::select(study_id, tempus_date, assay_type, tumor_purity)
mat <- mat %>% filter(
    (assay_type == "Tempus xF" & tempus_date == "6/18/25") |
    (assay_type == "Tempus xT" & tempus_date == "1/4/23")
  )

#Filter for Tempus reports and convert to datetimes
vaf_mat <- mat %>% filter(
    (assay_type == "Tempus xF" & tempus_date == "6/18/25") |
    (assay_type == "Tempus xT" & tempus_date == "1/4/23")
  ) %>% mutate(Date = as.Date(
  tempus_date,
  format = "%m/%d/%y")) %>% arrange(Date)%>% dplyr::select(-c(pluvicto_start, pluvicto_end, tempus_date, assay_type, Date))  %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>% t()

#Distance of VAFs using H clustering
d <- dist(vaf_mat)
hc <- hclust(d, method = "average")

#Manually inspect clustering to determine cutoff
plot(hc)
# cut tree by distance threshold
groups <- cutree(hc, h = 20)  # tune h

#Rename timepoints
cnames <-c( paste0("T", seq_len(ncol(vaf_mat))), "groups")
vaf_mat <- data.frame(vaf_mat, groups) 
colnames(vaf_mat) <- cnames
#Average clusters in same grouping based on clustering and concatenate labels to get clones
#*We assume that VAF is proportional to CCF although no CN data to determine this**
vaf_mat <- vaf_mat %>% mutate(mutation = row.names(.)) %>% 
 group_by(groups) %>%
  summarize(
    across(where(is.numeric), mean),
    mutations = paste(mutation, collapse = "\n"),
    .groups = "drop"
  )  %>%
  dplyr::select(-groups) %>%                 # drop grouping column
  column_to_rownames("mutations") %>% # set clone names
  as.matrix()

#Manually inspect VAF trajectories to determine parent child relationships, more than one solution may be possible
data.frame(vaf_mat)

#xT Purity from Tempus reports
purity <- c(T1 = 60, T2 = 22.2)
mat_corrected <- sweep(vaf_mat, 2, purity/100, "/")
mat_corrected <- pmin(mat_corrected, 100)#Pin max at 100% CCF
data.frame(mat_corrected)

```


```{r}
parents <- c(0, 0, 2, 0)
timepoints <- c(1, 2)


#Get tempus and pluvicto dates
tempus_dates <- mdy(mat$tempus_date) %>% sort()
pluvicto_dates <- mdy(unique(c(mat$pluvicto_start, mat$pluvicto_end))) %>% sort()
t0 <- min(tempus_dates)

tempus_time_numeric <- as.numeric(tempus_dates - t0)
pluvicto_time_numeric <- as.numeric(pluvicto_dates - t0)

#Create annotations for lines of times and labels
all_vlines <- c(
  tempus_time_numeric,
  pluvicto_time_numeric)

all_vlabels <- c(
  format(tempus_dates, "%b %Y"),
  paste0("PRPT\n", format(pluvicto_dates, "%b %Y")))

#Grep CN data
cn_df <- data %>%
  dplyr::filter(study_id == "WU-LU-066" & assay_type == "Tempus xT") %>%
  mutate(
    Tempus_date = mdy(tempus_date)
  ) %>%
  dplyr::select(
    study_id,
    Tempus_date,
    mut_status
  ) %>%
  mutate(
    cn_events = str_extract_all(
      mut_status,
      regex(
         "(\\b[A-Z0-9]+\\b)\\s*(CN gain|copy number gain|amplification|amp|CN deletion|copy number loss|loss|del|copy_number_gain|copy_number_loss|CopyNumberLoss)",
        ignore_case = TRUE
      )
    )
  ) %>%
  unnest(cn_events) %>%
  mutate(
    gene = str_extract(cn_events, "^[A-Z0-9]+"),
    cn_type = case_when(
      str_detect(cn_events, regex("gain|amp", ignore_case = TRUE)) ~ "gain",
      str_detect(cn_events, regex("loss|del", ignore_case = TRUE)) ~ "loss"
    )
  )%>% 
  mutate(
    time_numeric = as.numeric(Tempus_date - t0)
  )

#CN event label associated with tempus reports for GOIs
cn_annotations <- cn_df %>% dplyr::filter(gene %in% c("TP53", "AR", "BRCA2", "ATM", "PTEN", "RB1")) %>% 
  group_by(time_numeric) %>%
  summarise(
    x   = dplyr::first(time_numeric),
    y   = 110,
    lab = paste(
      paste0(gene, " ", cn_type),
      collapse = ";"
    ),
    col = c("black"),
    .groups = "drop"
  )

#Create Seaobject
seaObject <- createSeaObject(round(vaf_mat, 2), parents, timepoints = tempus_time_numeric,
                            timepointInterpolation = T, col = pal_jco(alpha = 0.5)(10)[1:4], cloneLabels = row.names(vaf_mat))

pdf("/Users/debanjansaha/Desktop/pluvicto_plots_v2/WU-LU-066_dolphinplot.pdf", width = 10)
#Create dolphin plot
dolphinPlot(
  seaObject,
  vlines = all_vlines,
  vlab   = all_vlabels,
  vlabSize = 4,
  showLegend = TRUE,
  markMeasuredTimepoints = tempus_time_numeric,annotations = cn_annotations,annotSize = 4,
  borderCol = "black",
  shape = "spline",
  pos = "center",
  separateIndependentClones = TRUE, ylab = "Variant allele frequency", xlab = "Time"
)
dev.off()

```

#WU-LU-032
```{r}
#Extract SNVs/Indels for annotated mutations and vus
mat <- data %>% dplyr::filter(study_id == "WU-LU-032") %>% mutate( mutations = paste0(mut_status, "|", vus), extracted = str_extract_all(mutations, vaf_regex))  %>%  unnest(extracted) %>%  mutate(
    gene    = str_match(extracted, vaf_regex)[,2],
    protein = str_match(extracted, vaf_regex)[,3],
    vaf     = as.numeric(str_match(extracted, vaf_regex)[,4])) %>% mutate(protein =gsub(pattern = "VGS", replacement = "", x = protein) ) %>% mutate(mut_id = paste0(gene, "_", protein)) %>% dplyr::select(pluvicto_start, pluvicto_end, , tempus_date, assay_type, mut_id, vaf) %>% pivot_wider(names_from = mut_id, values_from = vaf)

#Filter for Tempus reports and convert to datetimes
vaf_mat <- mat %>% dplyr::filter(grepl(pattern = "xF|xT", x = assay_type)) %>% mutate(Date = as.Date(
  tempus_date,
  format = "%m/%d/%y")) %>% arrange(Date)%>% dplyr::select(-c(pluvicto_start, pluvicto_end, tempus_date, assay_type, Date))  %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>% t()

#Distance of VAFs using H clustering
d <- dist(vaf_mat)
hc <- hclust(d, method = "average")

#Manually inspect clustering to determine cutoff
plot(hc)
# cut tree by distance threshold
groups <- cutree(hc, h = 10)  # tune h

#Rename timepoints
cnames <-c( paste0("T", seq_len(ncol(vaf_mat))), "groups")
vaf_mat <- data.frame(vaf_mat, groups) 
colnames(vaf_mat) <- cnames
#Average clusters in same grouping based on clustering and concatenate labels to get clones
#*We assume that VAF is proportional to CCF although no CN data to determine this**
vaf_mat <- vaf_mat %>% mutate(mutation = row.names(.)) %>% 
 group_by(groups) %>%
  summarize(
    across(where(is.numeric), mean),
    mutations = paste(mutation, collapse = "\n"),
    .groups = "drop"
  )  %>%
  dplyr::select(-groups) %>%                 # drop grouping column
  column_to_rownames("mutations") %>% # set clone names
  as.matrix()

#Manually inspect VAF trajectories to determine parent child relationships, more than one solution may be possible
data.frame(vaf_mat)

#Scraped from Tempus report in EPIC, if reported as median VAF in cfDNA, then Purity = 2 * VAF
#T1: 90% (xT)
#T2: 1.8% (xF)
#T3: 10.6% (xF)
purity <- c(T1 = 90, T2 = 2 * 1.8, T3 = 2 * 10.6)
mat_corrected <- sweep(vaf_mat, 2, purity/100, "/")

mat_corrected <- pmin(mat_corrected, 100)#Pin max to 100% CCF
data.frame(mat_corrected)

```


```{r}
parents <- c(0, 1, 1, 1)
timepoints <- c(1, 2, 3)


#Get tempus and pluvicto dates
tempus_dates <- mdy(mat$tempus_date) %>% sort()
pluvicto_dates <- mdy(unique(c(mat$pluvicto_start, mat$pluvicto_end))) %>% sort()
t0 <- min(tempus_dates)

tempus_time_numeric <- as.numeric(tempus_dates - t0)
pluvicto_time_numeric <- as.numeric(pluvicto_dates - t0)

#Create annotations for lines of times and labels
all_vlines <- c(
  tempus_time_numeric,
  pluvicto_time_numeric)

all_vlabels <- c(
  format(tempus_dates, "%b %Y"),
  paste0("PRPT\n", format(pluvicto_dates, "%b %Y")))

#Grep CN data
cn_df <- data %>%
  dplyr::filter(study_id == "107457016" & assay_type == "Tempus xT") %>%
  mutate(
    Tempus_date = mdy(tempus_date)
  ) %>%
  dplyr::select(
    study_id,
    Tempus_date,
    mut_status
  ) %>%
  mutate(
    cn_events = str_extract_all(
      mut_status,
      regex(
         "(\\b[A-Z0-9]+\\b)\\s*(CN gain|copy number gain|amplification|amp|CN deletion|copy number loss|loss|del|copy_number_gain|copy_number_loss|CopyNumberLoss)",
        ignore_case = TRUE
      )
    )
  ) %>%
  unnest(cn_events) %>%
  mutate(
    gene = str_extract(cn_events, "^[A-Z0-9]+"),
    cn_type = case_when(
      str_detect(cn_events, regex("gain|amp", ignore_case = TRUE)) ~ "gain",
      str_detect(cn_events, regex("loss|del", ignore_case = TRUE)) ~ "loss"
    )
  )%>% 
  mutate(
    time_numeric = as.numeric(Tempus_date - t0)
  )

#CN event label associated with tempus reports for GOIs
cn_annotations <- cn_df %>% dplyr::filter(gene %in% c("TP53", "AR", "BRCA2", "ATM", "PTEN", "RB1")) %>% 
  group_by(time_numeric) %>%
  summarise(
    x   = dplyr::first(time_numeric),
    y   = 110,
    lab = paste(
      paste0(gene, " ", cn_type),
      collapse = ":"
    ),
    col = c("black"),
    .groups = "drop"
  )

#Create Seaobject
seaObject <- createSeaObject(round(mat_corrected, 2), parents, timepoints = tempus_time_numeric,
                            timepointInterpolation = T, col = pal_jco(alpha = 0.5)(10)[1:4], cloneLabels = row.names(vaf_mat))
pdf("/Users/debanjansaha/Desktop/pluvicto_plots_v2/WU-LU-032_dolphinplot.pdf", width = 15, height = 12)
#Create dolphin plot
dolphinPlot(
  seaObject,
  vlines = all_vlines,
  vlab   = all_vlabels,
  vlabSize = 5,
  showLegend = TRUE,
  markMeasuredTimepoints = tempus_time_numeric,annotations = cn_annotations,annotSize = 10,
  borderCol = "black",
  shape = "spline",
  pos = "center",
  separateIndependentClones = TRUE, ylab = "Variant allele frequency", xlab = "Time"
)
dev.off()
```


########################KM plots######################################
#OS analysis in Pluvicto treated patients regressed onto mutation status
```{r}
library(survival)


#today <- Sys.Date()
today <- as.Date("2026-5-1")
#2026-5-1


```


```{r}
# Reusable helper: does mut_status show a mutation (any type) OR the specified CNA direction for `gene`?
gene_altered <- function(mut_status, gene, cna_keywords) {
  pattern <- paste0(
    "\\b", gene, "\\b\\s*(",
      "p\\.[A-Za-z0-9_*]+",     # missense/nonsense/other HGVSp
      "|splice_region",         # matches splice_region, splice_region_variant, etc.
      "|fs\\b",
    ")",
    "|\\b", gene, "\\b.*(", cna_keywords, ")"
  )
  as.integer(coalesce(str_detect(mut_status, regex(pattern, ignore_case = TRUE)), FALSE))
}

cna_loss <- "CN loss|copy number loss|deletion|del|copy_number_loss|CopyNumberLoss"
cna_gain <- "CN gain|copy number gain|amplification|amp|copy_number_gain"

#Subset dataframe for patient's with AR, TP53, BRCA2, ATM, or PTEN mutations
df <- data %>%
  dplyr::filter(!is.na(pluvicto_start)) %>%
  mutate(
    pluvicto_start = mdy(pluvicto_start),
    death_date     = mdy(death_date),
    Tempus_date    = mdy(tempus_date),
    status = if_else(!is.na(death_date), 1L, 0L),
    followup_date = if_else(status == 1L, death_date, today),
    os_time_days = as.numeric(followup_date - pluvicto_start),
    os_time_months = os_time_days / 30,

    AR_mut    = gene_altered(mut_status, "AR",    cna_gain),
    TP53_mut  = gene_altered(mut_status, "TP53",  cna_loss),
    BRCA2_mut = gene_altered(mut_status, "BRCA2", cna_loss),
    ATM_mut   = gene_altered(mut_status, "ATM",   cna_loss),
    PTEN_mut  = gene_altered(mut_status, "PTEN",  cna_loss)
  )
```


```{r}
#Filter for entries with Tempus reports prior to Pluvicto start and not induced by Pluvicto, dedup patient entries
#Indicator variable if any mutation (annotated or VUS) at any VAF was detected in the dataframe 
df_patient <- df %>%
  dplyr::filter(Tempus_date <= pluvicto_start) %>%
  group_by(study_id) %>%
  summarise(
    AR_mut = as.integer(any(AR_mut == 1, na.rm = TRUE)),
    PTEN_mut = as.integer(any(PTEN_mut == 1, na.rm = TRUE)),
    TP53_mut = as.integer(any(TP53_mut == 1, na.rm = TRUE)),
    BRCA2_mut = as.integer(any(BRCA2_mut == 1, na.rm = TRUE)),
    ATM_mut = as.integer(any(ATM_mut == 1, na.rm = TRUE)),

    pluvicto_start = dplyr::first(pluvicto_start),
    death_date = dplyr::first(death_date),
    last_followup = dplyr::first(followup_date),
    status = dplyr::first(status),
    os_time_months = dplyr::first(os_time_months),

    .groups = "drop"
  )

#Add PSA50 grouping variable 
df_patient <- df_patient %>%
  mutate(group = case_when(
    study_id %in% psa50_pts    ~ "PSA50 Responders",
    study_id %in% nonpsa50_pts ~ "Non-Responders",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(group))

#View(df_patient)
```


```{r}

#Set theme
pub_theme <- theme_classic() +
  theme(
    axis.text       = element_text(color = "black", size = 11),
    axis.title      = element_text(color = "black", size = 12),
    plot.title      = element_text(hjust = 0.5, size = 13, face = "bold"),
    legend.text     = element_text(position = "None"),
    legend.title    =element_text(position = "None"), 
    panel.border    = element_rect(color = "black", fill = NA)
  )

#Mutations vector
mutations <- c("TP53_mut", "PTEN_mut", "AR_mut", "BRCA2_mut", "ATM_mut")
mut_labels <- c("TP53", "PTEN", "AR", "BRCA2", "ATM")

#Loop to create KM plots
km_plots <- lapply(seq_along(mutations), function(i) {
  mut   <- mutations[i]
  label <- mut_labels[i]

  #Temp Df to create 4 category groupings 
  df_tmp <- df_patient %>%
    mutate(strata = case_when(
      .data[[mut]] == 0 & group == "Non-Responders"   ~ "WT | Non-Responder",
      .data[[mut]] == 0 & group == "PSA50 Responders" ~ "WT | PSA50 Responder",
      .data[[mut]] == 1 & group == "Non-Responders"   ~ "Altered | Non-Responder",
      .data[[mut]] == 1 & group == "PSA50 Responders" ~ "Altered | PSA50 Responder"
    ))

  #Create Survfit object
  fit <- survfit(Surv(os_time_months, status) ~ strata, data = df_tmp)

  #Plotting all four curves
  ggsurvplot(
    fit,
    data              = df_tmp,
    conf.int          = FALSE,
    pval              = TRUE,
    pval.size         = 4,
    risk.table        = TRUE,
    risk.table.col    = "strata",
    risk.table.height = 0.28,
    fontsize          = 3.5,
    title             = label,
    xlab              = "Months since Pluvicto start",
    ylab              = "OS (%)",
    break.time.by     = 6,
    fun               = function(x) x * 100,
    palette           = pal_jco()(4),
    censor.shape      = 124,
    censor.size       = 4,
    ggtheme           = pub_theme,
    tables.theme      = theme_cleantable()
  )
})

names(km_plots) <- mut_labels

#Cox PH modeling for univariate analyses 
coxph(Surv(os_time_months, status) ~ AR_mut + group + AR_mut*group, data = df_patient) %>% summary()
coxph(Surv(os_time_months, status) ~ TP53_mut + group + TP53_mut*group, data = df_patient) %>% summary()
coxph(Surv(os_time_months, status) ~ ATM_mut + group + ATM_mut*group, data = df_patient) %>% summary()
coxph(Surv(os_time_months, status) ~ PTEN_mut + group + PTEN_mut*group, data = df_patient) %>% summary()
coxph(Surv(os_time_months, status) ~ BRCA2_mut + group + BRCA2_mut*group, data = df_patient) %>% summary()


#Median OS across all patients
df_patient$os_time_months %>% summary()


#Plotting
km_plots_no_legend <- lapply(km_plots, function(p) {
  p$plot <- p$plot + theme(legend.position = "none")
  p
})

pdf("KM_plots_pluvicto_mut_psa50.pdf", width = 15, height = 10)
arrange_ggsurvplots(
  km_plots_no_legend, 
  ncol = 3,
  nrow = 2,
  risk.table.height = 0.25
)
dev.off()

```

```{r}
#Create COX PH model per mutation
genes <- c("ATM_mut")

fits <- lapply(genes, function(g) {
  coxph(as.formula(paste("Surv(os_time_months, status) ~", g, "* group")), data = df_patient)
})
names(fits) <- genes

#Getter for summary stats, HR, CI, P value
forest_df <- lapply(names(fits), function(gene) {
  s <- summary(fits[[gene]])

  #Df of stats
  coef_df <- as.data.frame(s$conf.int)
  p_vals  <- s$coefficients[, "Pr(>|z|)"]
  
  data.frame(
    gene  = gene,
    term  = rownames(s$coefficients),
    hr    = coef_df[, "exp(coef)"],
    lower = coef_df[, "lower .95"],
    upper = coef_df[, "upper .95"],
    pval  = p_vals
  )
}) %>%
  bind_rows() %>%
  mutate(

      #Str replace to clean vars    
      clean_label = term %>%
      str_replace_all(c("group" = "", "ATM_mut" = "ATM", "AR_mut" = "AR")),
    sig_stars = case_when(
      pval < 0.001 ~ "***",
      pval < 0.01  ~ "**",
      pval < 0.05  ~ "*",
      TRUE         ~ ""
    ),
    display_name = paste0(gene, ": ", clean_label)
  )

#Plotting
p2 <- forest_df %>%
  mutate(
    term_clean = term %>% 
      str_replace_all("groupPSA50 Responders", "PSA50 Response") %>%
      str_replace_all("ATM_mut", "Mutation Effect") %>%
      str_replace_all("AR_mut", "Mutation Effect"),
    term_clean = str_replace(term_clean, ":", " x "),
    
    stat_label = sprintf("HR: %.2f (p=%.3f)", hr, pval)
  ) %>% 
  ggplot(aes(x = hr, y = term_clean)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2, color = viridis(1, begin = 0.2)) +
  geom_point(size = 3, color = viridis(1, begin = 0.1)) +
  geom_text(aes(label = stat_label), 
            hjust = -0.1,  
            vjust = -1,    #
            size = 3.5, 
            color = "black") +
  facet_wrap(~gene, ncol = 1, scales = "free_y") + 
  scale_x_log10(expand = expansion(mult = c(0.1, 0.4))) + 
  theme_bw(base_size = 20) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    panel.spacing = unit(1.5, "lines")
  ) +
  labs(
    title = "Cox PH model of OS\nwith ATM mutation and PSA50",
    x = "Hazard Ratio (Log Scale)",
    y = ""
  )
p2

#ggsave(plot = p2, filename = "ATM_PSA50_forest.pdf", width = 10)


```





#PSA50 analysis with mutations.##################################### 
```{r}
#Pull PSA value columns and pivot long
psa_heat <- data %>%
  dplyr::select(study_id, 
                `Pre-C1` = psa_pre_c1_value,
                `C1`     = psa_c1_value,
                `C2`     = psa_c2_value,
                `C3`     = psa_c3_value,
                `C4`     = psa_c4_value,
                `C5`     = psa_c5_value,
                `C6`     = psa_c6_value) %>%
  distinct(study_id, .keep_all = TRUE) %>%
  mutate(across(-study_id, as.numeric)) %>%
  #Absolute difference columns
  mutate(across(c(`C1`, `C2`, `C3`, `C4`, `C5`, `C6`),
                ~ . - `Pre-C1`,
                .names = "diff_{.col}")) %>%
  #Percent change columns
  mutate(across(c(`C1`, `C2`, `C3`, `C4`, `C5`, `C6`),
                ~ (. - `Pre-C1`) / `Pre-C1` * 100,
                .names = "pct_{.col}")) %>%
  mutate(`Pre-C1` = 0) %>%
  dplyr::select(study_id, starts_with("diff_"), starts_with("pct_")) %>%
  pivot_longer(
    cols      = -study_id,
    names_to  = c(".value", "Cycle"),
    names_sep = "_"
  ) %>%
  mutate(Cycle = factor(Cycle, levels = c("Pre-C1","C1","C2","C3","C4","C5","C6"))) %>%
  group_by(study_id) %>%
  mutate(PSA50 = any(pct <= -50, na.rm = TRUE)) %>%
  ungroup()
```

#Filtering for PSA50 analysis (67 total patients with PCa and Tempus data --> 37 recieved Pluvicto --> 33 had > 1 PSA (i.e 4 pts never had follow-up after Pluvicto) --> 15 achieved PSA50 and 18 did not) 
```{r}
#Filtering for patient's who had recurrent PSA values (>1 PSA) 
#psa_heat %>% filter(study_id %in% got_pluvicto$study_id) %>% .$study_id %>% unique()


psa_data <- psa_heat %>% group_by(study_id) %>% filter(study_id %in% got_pluvicto$study_id) %>% 
  filter(n() > 1 & any(!is.na(pct))) %>%
  ungroup() %>% .$study_id %>% unique() %>% data.frame() %>% .$.
psa_heat <-  psa_heat %>% filter(study_id %in% psa_data)
#Filter for patients who had PSA response
#Variable denoting pateint's who had PSA50 at any point in treatment cycle (i.e PSA decline of 50% from baseline pre cycle 1 PSA)
psa50_pts <- psa_heat %>% filter(PSA50 == T) %>% .$study_id %>% unique() 
nonpsa50_pts <- psa_heat %>% filter(PSA50 == F) %>% .$study_id %>% unique() 
psa_heat %>% group_by(study_id) %>% unique()
```

#PSA kinetics plot for trajectory for 33 patients out of 37 
```{r}
#pdf("psa_plot.pdf", width = 10, height = 10)
p1 <- psa_heat %>%
  mutate(Cycle = factor(Cycle, levels = c("Pre-C1","C1","C2","C3","C4","C5","C6"))) %>%
  ggplot(aes(x = Cycle, y = pct, color = PSA50, group = study_id)) +
  geom_point(size = 2) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  scale_color_jco( 
                        name = "PSA50\nduring treatemnt", 
                    
                        labels = c("TRUE" = "Responders", "FALSE" = "Non-Responders")) +
  theme_bw(base_size = 15) +
  labs(x = "Cycle", y = "%PSA change from Pre-C1 (ng/mL)", title = "PSA Trajectory by Cycle for 33 patients with > 1 PSA") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylim(-200, 900)
#dev.off()

#Dataframe from OS analysis of top 5 recurrently mutation genes (all SNVs/Indels/CNA; both annotated and VUS) and association with 
#PSA50 during RPLT

#Aggreate all mutations from OS analysis and group into responders/non-responders based on PSA 50
df_long <- df %>%
  dplyr::select(study_id, ends_with("_mut")) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders")) %>%
  pivot_longer(
    cols      = ends_with("_mut"),
    names_to  = "mutation",
    values_to = "has_mutation"
  ) %>%
  mutate(mutation = str_remove(mutation, "_mut"))

#Fisher exact testing for enrichment, followed by FDR correction for MHT
fisher_results <- df_long %>%
  group_by(mutation) %>%
  summarise(
    p_adj = p.adjust(fisher.test(table(has_mutation, group))$p.value, method = "BH"),
    sig   = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    ),
    .groups = "drop"
  )

#Summary statistics for count distribution of mutations
df_summary <- df_long %>%
  group_by(group, mutation) %>%
  summarise(count = sum(has_mutation, na.rm = TRUE), .groups = "drop") %>%
  left_join(fisher_results, by = "mutation") %>%
  mutate(mutation = fct_reorder(mutation, p_adj))

#Plotting
label_positions <- df_summary %>%
  group_by(mutation) %>%
  summarise(y_pos = max(count) + 1, sig = first(sig), .groups = "drop")

p2 <- ggplot(df_summary, aes(x = mutation, y = count, fill = group)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.7),
    vjust = -0.4, size = 3.2
  ) +
  geom_text(
    data = label_positions,
    aes(x = mutation, y = y_pos, label = sig),
    inherit.aes = FALSE, size = 5, fontface = "bold"
  ) +
  scale_fill_manual(values = c("Responders" =  jco_cols[5], "Non-Responders" =  jco_cols[6])) +
  scale_y_continuous(
    breaks = scales::pretty_breaks(),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Mutation Enrichment by PSA50 Response",
    subtitle = "Ordered by FDR-adjusted p-value | * p<0.05  ** p<0.01  *** p<0.001",
    x        = NULL, y = "Count (SNV/Indel/CNA)", fill = "Group"
  ) +
  theme_bw(base_size = 15) +
  theme(
    legend.position = "top",
    axis.text.x     = element_text(angle = 45, hjust = 1, face = "bold")
  )

pdf(file = "/Users/debanjansaha/Desktop/pluvicto_plots_v2/psa50.pdf", width = 15)
ggarrange(p1, p2, ncol = 2, nrow = 1, legend = T, common.legend = F)
dev.off()
```


```{r}
#Overall PSA50 rate
psa_heat$PSA50 %>% table() %>% prop.table()

df_long
psa_heat %>% inner_join(dplyr::select(df, c(study_id, ends_with("_mut"), starts_with("os")))) %>% View()

df_long %>%
  group_by(study_id, group) %>%
  summarise(has_mutation = max(has_mutation), .groups = "drop") %>%
  group_by(study_id, group) %>%
  summarise(has_mutation = max(has_mutation), .groups = "drop") %>%
  count(group, has_mutation)


15 + 22 = 37 - 3 = 34
```

#MSI/TMB/CH association with PSA50
```{r}
  
msi <- df %>%
  mutate(
    CH_mut = if_else(
      gene_altered(mut_status, "DNMT3A", cna_gain) == 1 |
      gene_altered(mut_status, "TET2",   cna_gain) == 1 |
      gene_altered(mut_status, "ASXL1",  cna_gain) == 1,
      1L, 0L
    ),
    tmb = as.numeric(tmb),
    MSI_high = if_else(str_detect(msi, regex("MSI-H|Instable|High", ignore_case = TRUE)), 1L, 0L),
    TMB_high = if_else(tmb >= 10, 1L, 0L)
  ) %>%
  group_by(study_id) %>%
  summarise(
    CH_mut   = as.integer(any(CH_mut == 1, na.rm = TRUE)),
    MSI_high = as.integer(any(MSI_high == 1, na.rm = TRUE)),
    TMB_high = as.integer(any(TMB_high == 1, na.rm = TRUE))
  ) %>% mutate(group = case_when(
    study_id %in% psa50_pts    ~ "PSA50 Responders",
    study_id %in% nonpsa50_pts ~ "Non-Responders",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(group))


vars <- c("CH_mut", "MSI_high", "TMB_high")

results <- map_dfr(vars, function(v) {
  tbl <- table(msi[[v]], msi$group)
  test <- fisher.test(tbl)
  tibble(
    variable = v,
    p_value  = test$p.value,
    odds_ratio = test$estimate
  )
})

results

```


#PSMA discordance with mutation/PSA50 status########################
```{r}
df_long

df$fdg_response 

df %>%
  dplyr::select(study_id, ends_with("_mut"), fdg_response) %>% filter(!is.na(fdg_response)) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders")) %>%
  pivot_longer(
    cols      = ends_with("_mut"),
    names_to  = "mutation",
    values_to = "has_mutation"
  ) %>% summarise(
    p_adj = p.adjust(fisher.test(table(group, fdg_response))$p.value, method = "BH"),
    sig   = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    ),
    .groups = "drop"
  )
```


```{r}
res <- df %>% 
  filter(!is.na(fdg_response) & grepl(pattern = "Discordance|Concordance", x = fdg_response)) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders")) %>%
  mutate(fdg = ifelse(grepl(pattern = "Concordance", x = fdg_response), yes = "Concordance", no = "Discordance")) %>% 
  count(group, fdg) %>%
  group_by(group) %>%
  mutate(pct = n / sum(n),
         total = sum(n)) %>%
  ungroup()


df %>% 
  filter(!is.na(fdg_response) & grepl(pattern = "Discordance|Concordance", x = fdg_response)) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders")) %>%
  mutate(fdg = ifelse(grepl(pattern = "Concordance", x = fdg_response), yes = "Concordance", no = "Discordance")) %>% 
  count(group, fdg)

p_val <- fisher.test(table(df %>%
  filter(!is.na(fdg_response)) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders")) %>%
  pull(group),
  df %>% filter(!is.na(fdg_response)) %>% pull(fdg_response)))$p.value

sig <- case_when(p_val < .001 ~ "***", p_val < .01 ~ "**", p_val < .05 ~ "*", TRUE ~ "ns")

ggplot(res, aes(x = group, y = pct, fill = fdg)) +
  geom_col(width = 0.6, color = "white") +
  geom_text(aes(label = paste0(round(pct*100), "%\n(", n, "/", total, ")")),
            position = position_stack(vjust = 0.5), size = 3, color = "white") +
  annotate("text", x = 1.5, y = 1.05, label = sig, fontface = "bold", size = 5) +
  scale_fill_manual(values = pal_aaas()(5)) +  # adjust to # of fdg_response levels
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = NULL, fill = "FDG response") +
  theme_minimal()
```
```{r}

res <- df %>%
  select(study_id, ends_with("_mut"), fdg_response) %>%  # drop this select line if not needed
  filter(!is.na(fdg_response) & grepl("Discordance|Concordance", fdg_response)) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders"),
         fdg   = ifelse(grepl("Concordance", fdg_response), "Concordance", "Discordance")) %>%
  pivot_longer(ends_with("_mut"), names_to = "mutation", values_to = "has_mutation") %>%
  filter(has_mutation == 1) %>%
  count(mutation, group, fdg) %>%
  group_by(mutation, group) %>%
  mutate(pct = n / sum(n), total = sum(n)) %>%
  ungroup()

# fisher p-value per mutation
sig_tab <- df %>%
  filter(!is.na(fdg_response) & grepl("Discordance|Concordance", fdg_response)) %>%
  mutate(group = ifelse(study_id %in% psa50_pts, "Responders", "Non-Responders"),
         fdg   = ifelse(grepl("Concordance", fdg_response), "Concordance", "Discordance")) %>%
  pivot_longer(ends_with("_mut"), names_to = "mutation", values_to = "has_mutation") %>%
  filter(has_mutation == 1) %>%
  group_by(mutation) %>%
  filter(n_distinct(group) == 2, n_distinct(fdg) == 2) %>%   # fisher needs both levels present
  summarise(p = fisher.test(table(group, fdg))$p.value, .groups = "drop") %>%
  mutate(p_adj = p.adjust(p, method = "BH"),
         sig = case_when(p_adj < .001 ~ "***", p_adj < .01 ~ "**", p_adj < .05 ~ "*", TRUE ~ "ns"))

ggplot(res, aes(x = group, y = pct, fill = fdg)) +
  geom_col(width = 0.6, color = "white") +
  geom_text(aes(label = paste0(round(pct*100), "%\n(", n, "/", total, ")")),
            position = position_stack(vjust = 0.5), size = 2.8, color = "white") +
  geom_text(data = sig_tab, aes(x = 1.5, y = 1.05, label = sig),
            inherit.aes = FALSE, fontface = "bold", size = 4) +
  facet_wrap(~ mutation) 


```

